---
title: "ds4pp\\Problem Set 2b"
author: "Namita Karnik"
date: "Jan. 31, 2024"
format:  
  html:
    embed-resources: true
---

```{r setup}
#| include: false
library(tidyverse)
```

Due to bcourses by the start of class (Thursday, Feb. 1 at 10:30am).

Name your submission files `ps2b.qmd` and `ps2b.html`.

# Plotting CA crime statistics

Starting next week, we'll be looking at California crime statistics from the California Open Justice Initiative. Our data set next week has a much more complex structure. But for now, let's start with some simple statistics aggregated to the county level. Both the `csv` file and the corresponding data dictionary are on bcourses.

First, let's load in the file. (*Hint:* Make sure the `csv` file is in the same directory as the `qmd` file and then set the working directory to the "source file location.")

```{r}
crime <- read_csv("ca_crime_stats.csv")
crime
```

1.  Plot a histogram of the total arrest rate per 100,000 people by county in 2014. Be sure to change the labels from the defaults.

```{r}
#| eval: false

crime |> 
  filter(year == 2014) |>
  ggplot(mapping = aes(x = arrests_total_rate_per_100000)) +
  geom_histogram() + 
  labs(x = "Arrests per 100,000 people"
       ,y = "Count"
       ,title = "Arrest rate per 100,000 by county in 2014") +
  theme_dark()
```

2.  Re-make this same plot as a density.

```{r}
crime |> 
  filter(year == 2014) |>
  ggplot(mapping = aes(x = arrests_total_rate_per_100000)) +
  geom_density() + 
  labs(x = "Arrests per 100,000 people"
       ,y = "Density"
       ,title = "Arrest rate per 100,000 by county in 2014")
```

3.  Compare the distributions of total arrest rate in 2005 and 2014 with overlapping densities.
Ans: In 2005, there were a higher number of average arrests, as compared to 2014. For example, as can be seen from this graph, in 2005 (red curve), there were a higher number of counties that arrested more than 6000 people per 100,000, than in 2014 (blue curve)
```{r}
crime |> 
  filter(year %in% c(2014,2005)) |> 
  ggplot(mapping = aes(x = arrests_total_rate_per_100000, fill = factor(year))) +
  geom_density() + 
  labs(
    x = "Arrests per 100,000 people",
    y = "Density",
    title = "Density of Arrest Rates per 100,000 by County for 2005 and 2014"
  )
```

4.  Re-make this comparison with boxplots.
Ans: I changed a few things while plotting a boxplot. I changed the number of arrests from x-axis (in the density plot), to the y axis. The trends however, remain the same. The total number of arrests per 100,000 people (across all counties) have fallen. 
```{r}
crime |> 
  filter(year %in% c(2014, 2005)) |> 
  ggplot(mapping = aes(x = factor(year), y = arrests_total_rate_per_100000, fill = factor(year))) +
  geom_boxplot() + 
  labs(
    x = "Year",
    y = "Arrest Rate per 100,000 people",
    title = "Comparison of Arrest Rates per 100,000 by County for 2005 and 2014")
```

5.  Make a scatterplot of the property crime rate and violent crime rate by county in 2014.

```{r}
crime |>
  filter(year == 2014) |>
  ggplot() +
  geom_jitter(mapping = aes(x = factor(county), y = violent_crimes + property_crimes), width = 0.2) + 
  labs(
    x = "Counties",
    y = "Crimes Rate (Violent and Property) by County in 2014",
    title = "Violent and Property crime rate by County in 2014"
  )
```

6.  Make a line plot of the total arrest rate in Alameda county between 2005 and 2014.
Ans: The arrests in Alameda county have decreased over the years.

I also added the scales part because the years were in decimal points for some reason. 
```{r}

library(scales)

crime |> 
  filter(county == "Alameda") |> 
  ggplot(mapping = aes(x = year, y = arrests_total)) +
  geom_line() + 
  labs(x = "Year",
    y = "Total arrests in Alameda county",
    title = "Total arrests in Alameda county") +
    scale_x_continuous(labels = label_number(accuracy = 1))
```

7.  The five largest counties (by `population_served`) are given in the R code. On the same plot, make line plots of the arrest rates for these five counties between 2005 and 2014.

```{r}
five_largest_counties <- c("Los Angeles", "Orange", "Riverside", "San Bernardino", "San Diego")

#install.packages("scales")
library(scales)

crime |> 
  filter(county %in% c("Los Angeles", "Orange", "Riverside", "San Bernardino", "San Diego")) |>
  ggplot(mapping = aes(x = year, y = arrests_total_rate_per_100000, color = county)) +
  geom_line() + 
  labs(x = "Year",
    y = "Arrest rate per 100,000 people in each county",
    title = "Arrest rates per 100,000 people in five largest counties") +
    scale_x_continuous(labels = label_number(accuracy = 1))
```

8.  (Bonus) Use the `facet` option in `ggplot` to display these line plots as a grid of plots.
Ans: I also added the scales part because the years were in decimal points for some reason. 
```{r}
crime |> 
  filter(county %in% c("Los Angeles", "Orange", "Riverside", "San Bernardino", "San Diego")) |>
  ggplot(mapping = aes(x = year, y = arrests_total_rate_per_100000, color = county)) +
  geom_line() + 
  labs(x = "Year",
    y = "Arrest rate per 100,000 people in each county",
    title = "Arrest rates per 100,000 people in five largest counties") +
  facet_wrap( ~county, nrow = 3) +
  scale_x_continuous(breaks = c(2005,2007,2010,2012, 2014), labels = c("2005", "2007", "2010", "2012", "2014"))
```
